---
title: "Homework 8"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Model covariance matrix for subject i

$$
G_i = \begin{pmatrix}
\sigma^2_I & \sigma^2_{IS}\\
\sigma^2_{IS} & \sigma^2_S
\end{pmatrix}\\
Z_i = \begin{pmatrix}
1 & 0\\
1 & 1\\
1 & 2
\end{pmatrix}\\
Z_i^t = \begin{pmatrix}
1 & 1&1\\
0 & 1&2\\
\end{pmatrix}\\
R_i = \begin{pmatrix}
\sigma^2_e & 0 & 0\\
0 & \sigma^2_e & 0\\
0 & 0 & \sigma^2_e
\end{pmatrix}\\
V_i = Var(Y_i) = Z_iG_iZ_i^t + R_i=\begin{pmatrix}
1 & 0\\
1 & 1\\
1 & 2
\end{pmatrix}\begin{pmatrix}
\sigma^2_I & \sigma^2_{IS}\\
\sigma^2_{IS} & \sigma^2_S
\end{pmatrix}\begin{pmatrix}
1 & 1&1\\
0 & 1&2\\
\end{pmatrix}+\begin{pmatrix}
\sigma^2_e & 0 & 0\\
0 & \sigma^2_e & 0\\
0 & 0 & \sigma^2_e
\end{pmatrix}\\
= \begin{pmatrix}
\sigma^2_I+\sigma^2_e & \sigma^2_{IS} + \sigma^2_I & 2\sigma^2_{IS} + \sigma^2_I\\
\sigma^2_{IS} + \sigma^2_I & 2\sigma^2_{IS}+\sigma^2_I+\sigma^2_S+\sigma^2_e & 3\sigma^2_{IS}+\sigma^2_I+2\sigma^2_S\\
2\sigma^2_{IS}+\sigma^2_I & 3\sigma^2_{IS}+\sigma^2_I+2\sigma^2_S & 4\sigma^2_{IS}+\sigma^2_I+4\sigma^2_S+\sigma^2_e
\end{pmatrix}
$$

### 2. R code
```{r eval=FALSE}
mod1 <- lme(y ~ time, random=~time|id,data=dat)
```

### 3. Is it possible for this structure to have covariance that decays as time between responses increases?

In order to show this, you compare the covariance for times 0 and 1, and for times 0 and 2. If there's more correlation between time 0 and 1 than there is between time 0 and 2, then cov(0,1) > cov(0,2). This turns out to be fairly easy to rearrange, and shows that there can be decay as time between measurements increases, as long as $\sigma^2_{IS}<0$

$$
\sigma^2_{IS}+ \sigma^2_I > 2\sigma^2_{IS} + \sigma^2_I\\
\sigma^2_{IS}>2\sigma^2_{IS}\\
0 > \sigma^2_{IS}
$$