---
title: "Consulting Homework 1"
author: "Tim Vigers"
date: "September 2, 2018"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# R Lab 1

## Ex. 1 & 2
Add a red horizontal line at mean(y1) and a blue horizontal line at mean(y2). Move the legend up on the graph so it looks better and add red and blue points in front of the text.

```{r}
# Code below copied from R Lab 1 document. 
# Create 3 vectors of length 8
x <- c(1,3,4,6,9,10,11,14)
y1 <- c(102,101,89,92,81,80,75,72)
y2 <- c(88,85,81,80,76,71,66,64)
### Basic x-y graph
plot(x, y1)
### Useful options to plot
### Documentation in Help menu
### ?par gives options
plot(x, y1, xlab='x label', ylab='y label', pch=19, type="b", cex=.7, col="red", lty=4, lwd=3)
### Adding things to a plot
points(x, y2, pch="circle", cex=.7, col="blue")
### Note that some points are outside the plot.  To fix, set axis limits
plot(x, y1, xlab='x label', ylab='y label', pch=19, type="b", cex=.7, col="red",
      lty=4, lwd=3, xlim=c(0, 15), ylim=c(60, 110))
points(x, y2, pch="circle", cex=.7, col="blue")
# Add a red line at mean(y1).
lines(x,rep(mean(y1),8),col = "red")
# Add a blue line at mean(y2).
lines(x,rep(mean(y2),8),col = "blue")
# Move the legend (deleted previous legend).
text(12, 105, "Red = female", adj=0)
text(12, 100, "Blue = male", adj=0)
# Add red point in front of the text.
points(11.5, 105, pch=19, type="b", cex=.7, col="red")
# Add blue point in front of the text.
points(11.5, 100, pch="circle", cex=.7, col="blue")
```
\pagebreak

# R Lab 2

## Ex. 1
Use rep and seq to create c(.2,.2,.4,.4,.6,.6,.2,.2,.4,.4,.6,.6,.2,.2,.4,.4,.6.,6)

```{r}
# Make the sequence we're aiming for.
goal <- c(.2,.2,.4,.4,.6,.6,.2,.2,.4,.4,.6,.6,.2,.2,.4,.4,.6,.6)
# Use seq() to create c(0.2, 0.4, 0.6).
seq <- seq(.2,.6,by = 0.2)
# Use rep() to repeat each element in seq twice.
seq <- rep(seq,each = 2)
# Use rep() again to repeat the whole list three times.
seq <- rep(seq,times = 3)
# Show the sequence
seq
# Double check that it matches the goal sequence.
seq == goal
```

## Ex. 2

```{r}
# Generate a sample of 10000 normal values with mean 10 and sd 4.
r.norm <- rnorm(n = 10000, mean = 10, sd = 4)
# Make a histogram of the values.
hist(r.norm,xlab = "Value")
# Check that the empirical mean and SD match those used to generate the sample.
mean(r.norm)
sd(r.norm)
# For X~N(10, 16), find Pr(X > 18).
pnorm(18,mean = 10,sd = 4,lower.tail = FALSE)
# For X~N(10, 16), find the value so that 97.5% of the distribution is less than that value.
qnorm(0.975,mean = 10,sd = 4)
# Make a smooth line graph of the N(10, 16) density.
x <- seq(-5,25, by = 0.1)
y <- dnorm(x, mean = 10, sd = 4)
plot(x, y, type="l", main = "Normal Distribution (10, 16)", ylab = "Frequency", xlab = "Value")
```

## Ex. 3

With a shape parameter k and a scale parameter s, for a Gamma distribution: $$\mu = ks = 10 \text{ and } \sigma^2 = ks^2 = 16$$ So: $$k = \frac{100}{16} \text{ and } s = \frac{16}{10}$$

```{r}
# Generate a sample of 10000 values from a Gamma distribution with mean 10 and sd 4.
r.gamma <- rgamma(n = 10000, shape = 25/4, scale = 8/5)
# Verify empirically that your sample mean and sd are close to 10 and 4.
mean(r.gamma)
sd(r.gamma)
```
\pagebreak

# R Lab 3

```{r, echo=FALSE}
x <- c(1,3,4,6,9,10,11,14)      
y1 <- c(102,101,89,92,81,80,75,72)
y2 <- c(88,85,81,80,76,71,66,64)
lab <- c("a","b","c","d","e","f","g","h")

xy.mat <- cbind(x, y1, y2)
```

## Ex. 1
Explain (1 sentence each) what each of these statements does:

a. x[c(3:7)-2]: This command selects elements (3-2), (4-2), (5-2), (6-2), and (7-2), (i.e. elements 1 - 5) from the vector x.
```{r}
x
x[c(3:7)-2]
```

b. x[c(3:7)]-2: This takes elements 3 - 7 of vector x, and subtracts 2 from each value.
```{r}
x[c(3:7)]
x[c(3:7)]-2
```

c. xy.mat[14, ]: THis takes row 14 and all columns of the matrix xy.mat (subsetting row 8 is shown below, since there are only 8 rows of our matrix).
```{r}
xy.mat[8, ]
```

d. xy.mat[c(2:4)]: This takes the values in column 1, rows 2, 3, and 4.
```{r}
xy.mat[c(2:4)]
```

e. cbind(x, lab): This takes the vectors x and lab, and turns them into a matrix where each vector is a column.
```{r}
cbind(x, lab)
```

## Ex. 2
Using the object xy.mat and not making any new assignments (don't use <-), graph y2 versus x, omitting the 4th row, and with axis labels "y2" and "x".
```{r}
xy.mat
plot(xy.mat[c(1:3,5:8),1], xy.mat[c(1:3,5:8),3], xlab = "x", ylab = "y2")
```

## Ex. 3
Suppose y1 and y2 are measures of an outcome at times 1 and 2 for 8 subjects. So cbind(y1,y2) is the 'wide' form of the dataset. Create the 'long' form of the dataset, with 16 rows and 3 columns: a column for y, one for time, and one for subject id.
```{r}
# Make an empty matrix.
mat <- matrix(NA, ncol=3, nrow=16)
# Name the columns.
colnames(mat) <- c("subjectid", "time", "outcome")
# Fill in subject ids (from vector x).
mat[,1] <- rep(x,times = 2)
# Fill in times 1 and 2.
mat[,2] <- c(rep(1,times = 8),rep(2,times = 8))
# Fill in outcomes data from y1 and y2.
mat[,3] <- c(y1,y2)
# Show the matrix.
print.data.frame(as.data.frame(mat),row.names = FALSE)
```