---
title: "Consulting Homework 10"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE}
library(knitr)
library(pwr)
```

##a. Sample size calculations.
###i. Assuming a known SD

```{r echo=FALSE}
# Make Jamie's data table
pilot <- as.data.frame(rbind(c(1,1.81,3.55,8.28,10.33,6.7,13.28),
                             c(2,1.77,5,8.33,11.28,9.77,12.14),
                             c(3,2.09,NA,7.57,10.57,13.38,13.42),
                             c(4,1.09,NA,7.99,12.4,10.95,9.83),
                             c(5,1.66,3.81,7.94,10.46,16.32,11.54),
                             c(6,1.91,4.33,5.4,7.18,4.74,9.66),
                             c(7,1.31,1.28,6.71,6.25,17.13,7.52),
                             c(8,1.28,3.13,7.04,7.54,8.68,12.36)))
colnames(pilot) <- c("ID","2mg","5mg","10mg","15mg","20mg","30mg")
sds <- sapply(pilot[,2:7], function(x) round(sd(x,na.rm = T),7))
pilot <- rbind(pilot,c("SD",sds))
s <- as.numeric(sds[1])
```

```{r echo=FALSE}
kable(pilot, caption = "Total absorbed zinc (AZ) by dose")
```

Using the SD for the 2 mg group, because both the low and high zinc content groups are expected to absorb < 2 mg per day (90% power, alpha = 0.05):
$$
n = \frac{(\sigma_1^2 + \sigma_2^2)(Z_{0.9}+Z_{0.975})^2}{(\text{detectable difference})^2}=\frac{(0.3500204^2+0.3500204^2)*(1.28+1.96)^2}{(0.6)^2}=\frac{2.572212}{0.36} = 7.145033
$$

So for 90% power to detect a difference in AZ of 0.6 mg/day, assuming known variation, Jamie will need 8 participants in each group. 
\pagebreak

###ii. Assuming an unknown SD

Use R's built-in power calculation function:
```{r}
power.t.test(delta = 0.6,sd = s,power = 0.9,type = "two.sample")
```

With an unknown variance, R recommends a sample size of 9 participants per group.

##b. Sample size summary
The standard deviation for AZ after a 2 mg dose of zinc was used as an estimate of within group variation.  This value is 0.35 mg/day.  Assuming an alpha level of 0.05 for a two-sided test comparing two groups, a sample size of 9 subjects per group provides 90% power to detect a difference in AZ between groups of 0.6 mg/day.  This is a clinically meaningful difference that we consider a realistic expectation of the intervention in this study.  Assuming a drop-out rate of 20%, we will recruit 11 subjects per group.  

##c. Statistical analysis summary
Primary outcome is total zinc absorption.  Statistical analyses will be carried out using R. A two sided t test with significance level 0.05 will be used to compare absorption between the high and low zinc content groups.