---
title: "Consulting Final Exam"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##a. Sample size calculations with pilot data
###i. Assuming known variance

```{r echo=FALSE, include=FALSE}
# Read in data.
bpdata <- read.csv("/Users/timvigers/Documents/School/UC Denver/Biostatistics/Statistical Consulting 1/Final Exam/BIOS6621exam.csv")
```

Using the SD of BP change from the pilot dataset (excluding two outliers):

```{r echo=FALSE,include=FALSE}
# Calculate changes
change <- bpdata$SBP0 - bpdata$SBP1
# Remove outliers
change <- change[-c(which.max(change),which.min(change))]
# Create sample size calculation variables
sigma <- sd(change)
z1 <- qnorm(0.9)
z2 <- qnorm(0.975)
dd <- 5
```

$$
n = \frac{\sigma^2(Z_{0.9}+Z_{0.975})^2}{(\text{detectable difference})^2}=\frac{8.263^2 (1.282+1.960)^2}{(5)^2}=\frac{716.669}{25}=28.667
$$
So, with 29 participants you would be able to detect a 5 mmHg difference with 90% power at the alpha = 0.05 level (assuming known variance).

###ii. Assuming an unknown variance

Using R's built-in power calculation function:
```{r}
power.t.test(delta = 5, sd = sigma, power = 0.9, type = "one.sample")
```

With an unknown variance, R recommends a sample size of 31 participants.

##a. Sample size calculations based on He, F. J., et al. (2009).
###i. Assuming known variance

To calculate SD from He et al.'s mean change and 95% CI:

```{r echo=FALSE,include=FALSE}
n <- (71+69+29)
sigma2 <- sqrt(n) * (2.2)/3.92
```

$$
\sqrt{n} * \frac{(\text{upper limit}-\text{lower limit})}{2*1.96} = \sqrt{(71+69+29)}*\frac{(-3.0-(-5.2))}{3.92} = 
$$

```{r echo=FALSE}
sigma2
```

So:
$$
n = \frac{\sigma^2(Z_{0.9}+Z_{0.975})^2}{(\text{detectable difference})^2}=\frac{7.296^2 (1.282+1.960)^2}{(5)^2}=\frac{559.494}{25}=22.380
$$

###ii. Assuming an unknown variance

Using R's built-in power calculation function:
```{r}
power.t.test(delta = 5, sd = sigma2, power = 0.9, type = "one.sample")
```

With an unknown variance, R recommends a sample size of 25 participants.

##b. Sample size summary
The standard deviation for AZ after a 2 mg dose of zinc was used as an estimate of within group variation.  This value is 0.35 mg/day.  Assuming an alpha level of 0.05 for a two-sided test comparing two groups, a sample size of 9 subjects per group provides 90% power to detect a difference in AZ between groups of 0.6 mg/day.  This is a clinically meaningful difference that we consider a realistic expectation of the intervention in this study.  Assuming a drop-out rate of 20%, we will recruit 11 subjects per group.  