---
title: "Homework 5"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

```{r}
# Read in data
hw5 <- read.csv("/Users/timvigers/Documents/School/UC Denver/Biostatistics/Biostatistical Methods 2/Homeworks/Homework 5/hw5.txt",sep = "")
```
 
## Model 1: Change-Score Model

```{r}
mod1 <- lm(delta_FEV1 ~ 1,data = hw5)
sm <- summary(mod1)
sm
# MSE
mean((sm$residuals^2))
```

### a. Model equation

$$
Y_{post_i}-Y_{pre_i} = \beta_0 + \epsilon =0.076 + \epsilon \text{ ~ N}(0,0.017)
$$

### b. Interpretation

The difference between pre- and post-bronchodilator FEV1 is significantly different from 0 (p = 0.000198). 

### c. Simple test

A paired t-test gives the same result:
```{r}
t.test(hw5$pre_FEV1,hw5$post_FEV1,paired = T)
```

## Model 2: Baseline-as-Covariate Model

```{r}
mod2 <- lm(post_FEV1 ~ pre_FEV1, data = hw5)
sm <- summary(mod2)
sm
# MSE
mean((sm$residuals^2))
```

### a. Model equation

# Remove error term when in expected value form?

$$
Y_{post_i} = \alpha_0 + \alpha_1 Y_{pre_i}+ \epsilon =0.096 + 0.988 Y_{pre_i}+ \epsilon \text{ ~ N}(0,0.017)
$$

### b. Interpretation

Pre-bronchodilator FEV1 is significantly associated with post-bronchodilator FEV1 (p < 2e-16).

## Model 3: Hybrid Model

```{r}
mod3 <- lm(delta_FEV1 ~ pre_FEV1, data = hw5)
sm <- summary(mod3)
sm
# MSE
mean((sm$residuals^2))
```

### a. Model equation

$$
Y_{post_i}- Y_{pre_i} = \gamma_0 + \gamma_1 Y_{pre_i} + \epsilon =\\
0.096 - 0.012 Y_{pre_i} + \epsilon \text{ ~ N}(0,0.017)
$$

### b.

Change in FEV1 is not significantly associated with pre-bronchodilator FEV1 (p = 0.570)

### c. 

The difference between pre- and post-bronchodilator FEV1 is significantly different from 0 when controlling for baseline FEV1 (p = 0.021).

## Show algebraically that Model 1 is nested within Model 2

$$
\text{Model 1: }E[\Delta_i] = E[Y_{post_i}-Y_{pre_i}] = E[Y_{post_i}]-E[Y_{pre_i}]=\beta_0\\
\text{Model 2: }E[Y_{post_i}] = \alpha_0 + \alpha_1 Y_{pre_i}\\
\text{Therefore: Model 1 = Model 2}-E[Y_{pre_i}]
$$

This shows that model 2 is similar to model 1, but allows for variation in baseline FEV1. So the change-score model assumes that baseline values won't be significantly associated with the second time point values and just tests the difference between the two, while the baseline-as-covariate model tests whether there is an association. 

## Show algebraically that Models 2 and 3 are equivalent

$$
\text{Model 2: }E[Y_{post_i}] = \alpha_0 + \alpha_1 Y_{pre_i}\\
\text{Model 3: }E[Y_{post_i}-Y_{pre_i}] = E[Y_{post_i}]-E[Y_{pre_i}] =\gamma_0 + \gamma_1 Y_{pre_i}\\
E[Y_{post_i}] = \gamma_0 + \gamma_1 Y_{pre_i} + E[Y_{pre_i}]
$$

This shows that the models are telling you the same thing, but that $\gamma_1$ is slightly different from $\alpha_1$ because of the $E[Y_{pre_i}]$ term. You can see this in the model summaries too, since they have the same intercept ($\alpha_0 = \gamma_0$). However, in model 2 $\alpha_1$ tells you that for every unit increase in baseline FEV1, post-bronchodilator FEV1 increases by 0.9876 while $\gamma_1$ says that for the same change in baseline, the change in FEV1 decreases by 0.012. This is the same information, just parameterized in different ways.

## Model 4: Long Format

###a. 

The intercept in this model is the average pre-bronchodilator FEV1.

### b. 

$\hat\beta_1$ in this model is equivalent to $\hat\beta_0$ in model, which was the average difference between the timepoints. This makes sense because in model 4, $\hat\beta_1$ is the average change in FEV1 when going from the first time point to the second, so they are giving you the same information. 

### c. 

The standard error for $\hat\beta_1$ in model 4 is higher, because the equation for this  model essentially has an additional error term. Instead of adding $\epsilon_i$ where i = subjects 1,...,n, you're adding $\epsilon_{ij}$ where indexes time one or time 2. This raises the model MSE and increases the SE calculation for the covariates. 

### d. Simple test

A repeated measures ANOVA would produce the same results in this case. 