---
title: 'Stats Theory 1: HW 1'
author: "Tim Vigers"
date: "8/30/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro to R information
Create a new R markdown file: File -> New File -> R Markdown

Note: basic math can be done in R as well. Example:

```{r basicmath}
(5 + 6) / (7 * 3)
```

Learning objectives: dbinom, choose, and a for loop

If you need help with a function, type in help("function")

## 2b

A particular cancer experiment requires at least two highly susceptible mice. What is the probability that an order of 10 mice from the breeding facility will suffice?

Binomial formula: p(x) = choose(n, x) p^x (1-p)^(n-x)

Binomial R function: dbinom(x, size=n, prob, log = FALSE)

```{r 2b}
1 - dbinom(size = 10, prob = 0.75, x = 0) -  dbinom(size = 10, prob = 0.75, x = 1)
```

## 2c

Suppose that a particular batch of 7 mice actually contains exactly 3 highly susceptible ones. If the cancer researcher chooses three mice at random from this batch for a particular experiment, what is the probability that at least two of the three mice chosen will be highly susceptible ones?

choose function in R: choose(n, k) with n as size and k an integer

```{r 2c}
1 - (choose(3,0) * choose(4,3)) / choose(7,3) - (choose(3,1) * choose(4,2)) / choose(7,3)
13/35
```

## 2d
Notice that we can partition the event space into four possible subsets: 

i) 0 out of the 3 mice are highly susceptible.
```{r 2di}
dbinom(size = 3, prob = 0.05, x = 2)
```

ii) 1 out of the 3 mice is highly susceptible.
```{r 2dii}
dbinom(size = 2, prob = 0.05, x = 1) * 0.15 + (0.05)^2 * 0.85
```
iii) 2 out of the 3 mice are highly susceptible.
```{r 2diii}
choose(2,1) * (0.15) * (0.05) * (0.85) + (0.15)^2 * (0.95)
```
iv) 3 out of the 3 mice are highly susceptible.
```{r 2div}
choose(3,2) * (0.15)^2 * (0.85)
```

For loop in R: for(var in seq) expr 

for some variable var in the sequence seq, do the expression expr
```{r 2d}
for(i in 0:3) print((choose(3,i) * choose(4, 3-i)) / choose(7,3))
4/35
18/35
12/35
1/35
```



