---
title: "C & B 3.7 - Using Newton's Method"
author: "Carter Sevick"
date: "October 10, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the course of this problem we arrived at a point where we needed to solve the following function, set equal to zero, for t (the original problem used lambda).

```{r}

prb <- function(t) {
  exp(-t)*(1+t)- 0.01
}

```

Newton's method$^1$ was selected to attempt a solution, and to make that work we also need the derivative of the function:
(yes, there is an R function for this, ...but that would be cheating!)

```{r}

dprb <- function(t){
  -t*exp(-t)
}

```

The process of Newton's method is encapsulated in the following function:

```{r}

newton <- function(f              # function name to be solved 
                , df              # name of a function that is the derivative of f
                , initial=1       # an initial value to start the algorithm, default set at 1
                , tol=0.000000001 # tolerance level for testing convergence to the solution
                , maxiter = 50    # maximum number of iterations to perform befor giving up on a solution
) {
  
  .f. <- get(f)
  environment(.f.)  <- environment()

  .df. <- get(df)
  environment(.df.) <- environment()

  prev <- initial
  for(i in 1:maxiter) {
    
    newVal <- prev - .f.(prev)/.df.(prev)
    diff <- newVal - prev
   
    test <- abs(diff)<= tol
    if (test==FALSE & i>=maxiter) {print("the maximum number of iterations has been reached with no solution.")}
    if (test==TRUE) {break}
    
    prev <- newVal
  }
  list(solution = newVal, iter = i, diff=diff,converged=test)
}

```

Application of the algorithm:
```{r}

newton('prb', 'dprb')

```

References:

1. Anton H, Bivens I, Davis S "Calculus, Early Transcendentals, 7th" John Wiley and Sons, 2002