---
title: 'Stats Theory I: HW 3'
author: "Mayla Boguslav"
date: "8/13/2018"
output: html_document
---

## Intro to R information
Create a new R markdown file: File -> New File -> R Markdown

Note: basic math can be done in R as well. Example:

```{r basicmath}
(5 + 6) / (7 * 3)
```
If you need help with a function, type in help("function")

Learning objectives: plot and lines,

Reviewing: for loops, curves, latex

Note: plot and lines vs. curves

1. Use curves if you have one function to plot (one equation)

2. If more than one equation to plot, then use plot and lines



## (C+B 2.5)
Use theorem 2.1.8 to find the pdf of Y in Example 2.1.2. Show that the same answer is obtained by differentiating the cdf given in (2.1.6).

```{r 9}
curve((sin(x))^2, 0, 7) #Note that this is a function: each x has only 1 value of y

##Note that the inverse is not a function: if you pass a vertical line over the graph, it will intersect in several places. A clue to use the transformation over a partition theorem (2.1.8)

x <- seq(0,1,0.1)
plot(x,sqrt(asin(x)), type="l",  ylim=c(-8, 8))
lines(x,-sqrt(asin(x)))

```


## (C+B 2.8)
In each of the following show that the given function is a cdf and find $F^{-1}(y)$
A. $$F_X(x) = \begin{cases} 
0, & \mbox{if } x <0 \\ 1-e^{-x}, & \mbox{ if }x\geq 0 \end{cases}$$

```{r 11a}
curve(1-exp(-x), 0,100)
```


B. $$F_X(x) = \begin{cases} 
{\frac{e^x}{2}}, & \mbox{if } x <0 \\ \frac{1}{2}, & \mbox{ if }0\leq x<1 \\ 1-(\frac{e^{-x}}{2}), & \mbox{ if }1  \leq x \end{cases}$$

```{r 11b}
##the sequences for where each equation applies
x1 <- seq(-10,10,1)
x2 <- seq(-10,0,0.1)
x3 <- seq(1, 10, 0.1)
x4 <- seq(0,1,0.1)

##the plots themselves for each sequence
plot(x1, 0*x1,type="p", ylim=c(-1,1), col="blue",ylab="CDF")
lines(x2,exp(x2)/2)
lines(x3,1-(exp(-x3)/2))
lines(x4,0*x4+1/2)
```

C. $$F_X(x) =\begin{cases} 
\frac{e^x}{4}, & \mbox{if } x<0  \\ 1-\frac{e^{-x}}{4}, & \mbox{ if }x \geq 0 \end{cases} $$
```{r 11c}
##the sequences for where each equation applies
x1 <- seq(-10,10,1)
x2 <- seq(-10,0,0.1)
x3 <- seq(0, 10, 0.1)

##the plots themselves for each sequence
plot(x1, 0*x1,type="p", ylim=c(-1,1), col="blue",ylab="CDF")
lines(x2,exp(x2)/4)
lines(x3,1-(exp(-x3)/4))
```

# Sam's code for generating exponential data from uniform
Generate exponential(2) data:  Generate 1000 U(0,1) random variables and use these values to generate exponentail(2) random variables.
```{r random}
  N<-1000
  lambda<-2
  y<- (-lambda*log(1-runif(N,0,1)))
  mytxt<-paste("Distribution of ",N," RVs from an exponential(",lambda,")","\nBlue Line = true distribution",sep="")
  hist(y,main=mytxt,freq=F,xlab="X",ylab="pdf",breaks=c(0:20))
  xx<-seq(0,16,.01)
  lines(xx,dexp(xx,rate=(1/lambda)),lwd=2,col="blue")  # Note in help(dexp) the mean is 1/rate.
  ```




