---
title: "BIOS 6660 Homework 3"
author: "TODO: Your name"
date: "TODO: Today's date"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nycflights13)
library(dplyr)
library(ggplot2)
```


## Introduction

TODO: write a short introduction to the nycflights13 package and include a link to its online documentation.



```{r load_data}
# Load the flights dataset
data(flights)
```


## Data summary

TODO: write a short explanation of what the next code chunk does.



```{r summarize_data}
# TODO: print the column names of flights


# TODO: print the number of observations in flights

```


## Top flight destinations

TODO: write a short explanation of what the next code chunk does.



```{r top_dest}
# TODO: Print a table of the 10 most common destination cities and the number of flights to each destination.
#       Your table should have two columns labeled "dest" and "num_flights", and 10 rows.
#       Sort the table in decreasing order by num_flights.
#       Accomplish this with a sequence of dplyr verbs joined by the pipe operator.



```


## Arrival delays

TODO: write a short explanation of what the next code chunk does.



```{r delays}
# TODO: Print a table of the mean arrival delay by carrier.
#       Your table should have two columns labeled "carrier" and "mean_arr_delay_min".
#       Sort the table in decreasing order by mean_arr_delay_min
#       Accomplish this with a sequence of dplyr verbs joined by the pipe operator.



```


## Number of daily flights

TODO: write a short explanation of what the next code chunk does.



```{r daily_flights}
plot_daily_flights <- function(mon) {
  # Creates a histogram of daily flights by day for one month
  # (Bar chart with one column for each day of the month, where
  # the bar height is the total number of flights on that day)
  #
  # Args:
  #   mon: Month number e.g. 12
  #
  # Returns:
  #   Plot object representing a histogram of flights per day
  
  num_days <- max(filter(flights, month == mon)$day) # Number of days in the month
  plt_data <- flights %>% filter(month == mon) # Flight data for the month
  
  # TODO: Complete the body of this function and return a value.
  #       Use qplot or ggplot to create a histogram of flights per day.
  #       Use the day variable in plt_data.
  #       Use num_days from above for the number of bins in the histogram.
  #       Add a title to the plot that includes the month number (hint: use the "paste" function).
  #       Return the plot object.
  
  

    
}
```


TODO: write a short explanation of what the next code chunk does.



```{r plots, echo = FALSE}
# TODO: Plot histograms of daily flights for the months of June and November.
# TODO: Change a chunk option for this chunk so that the plots are included in the HTML
#       report, but the code in this chunk is not printed in the report.



```

