---
title: "Passing in the NBA"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "/Users/timvigers/Documents/GitHub/School/Analysis of Longitudinal Data/Final Project")
source("/Users/timvigers/Documents/GitHub/Tim-and-Laura/tim_r_functions.R")
library(knitr)
library(segmented)
library(nlme)
library(lme4)
library(ggrepel)
library(tidyverse)
```

```{r scrape,echo=FALSE,include=FALSE,cache=TRUE}
# Get data from basketball reference
source("./Code/scrape.R")
```

```{r data,echo=FALSE,include=FALSE}
# Read in passing CSV
passing <- read.csv("./Data/passing_data.csv")
# Colors
source("./Code/colors.R")
# Season to starting year
all_seasons$Season <- sub("-.*","",all_seasons$Season)
# Convert column types - all seasons
numcols <- c("Season","AST","FG","FGA","FG%","3P","3PA","3P%","STL",
             "BLK","TOV","PF","PTS")
all_seasons[,numcols] <- lapply(all_seasons[,numcols], 
                                function(x) as.numeric(as.character(x)))
# Team and season
all_seasons$team_season <- paste(all_seasons$Team,all_seasons$Season)
# Convert column types - passing
passing$Season <- sub("-.*","",passing$Season)
numcols <- c("Season","Passes.Made","Min")
passing[,numcols] <- lapply(passing[,numcols], 
                                function(x) as.numeric(as.character(x)))
```

## Questions

- Has the passing rate increased in the NBA?

- Do assists correlate with winning percentage?

# Passing

## Passes made per season

```{r pass plot,echo=FALSE,warning=FALSE,message=FALSE}
# Raw assist number plot
pass_plot <- 
  ggplot(passing,aes(x = Season,y = Passes.Made,group = Team)) + 
  geom_line(aes(color = Team)) + theme_bw() +
  scale_color_manual(values = primary_colors) + 
  ylab("Passes Made")
pass_plot
```

## Passing model selection

- Random intercept (RI) for team only:

lme(Passes.Made ~ Season, random = ~1|Team,
    data = passing,method = "ML")

- RI for team and random slope (RS) across season:

lme(Passes.Made ~ Season, random = ~Season|Team,
    data = passing,method = "ML")

- RI for team with AR(1) structure for repeated measures:

lme(Passes.Made ~ Season, random = ~1|Team,data = passing,
    correlation = corAR1(),method = "ML")

- RI for team and RS across season with AR(1) structure for repeated measures:

lme(Passes.Made ~ Season, random = ~Season|Team,data = passing,
    correlation = corAR1(),method = "ML")

## AIC

```{r aic,echo=FALSE}
passing_mod <- lme(Passes.Made ~ Season, random = ~1|Team,
                   data = passing,method = "ML")
passing_mod_ar1 <- lme(Passes.Made ~ Season, random = ~1|Team,
                       data = passing,correlation = corAR1(),method = "ML")
passing_mod_ar1_s <- lme(Passes.Made ~ Season, random = ~Season|Team,
                       data = passing,correlation = corAR1(),method = "ML")
aics <- as.data.frame(AIC(passing_mod,passing_mod_ar1,passing_mod_ar1_s))
rownames(aics) <- c("RI Only","RI and AR(1)","RI, RS, and AR(1)")
kable(aics)
```

The best model by AIC was random intercept for team with AR(1) structure for repeated measures (RI and RS did not converge without AR(1)). Minutes played did not affect the results.

## Passing results

There were no polynomial effects for time:

```{r, echo=FALSE}
passing_mod_ar1_poly <- lme(Passes.Made ~ poly(Season,4), random = ~1|Team,
                       data = passing,correlation = corAR1())
format_nlme_out(passing_mod_ar1_poly,caption = NULL)
```

## Break point

The segmented package in R suggests there's a knot at 2015:

```{r breakpoint}
linmod <- lm(Passes.Made ~ Season,data = passing)
segmented(linmod)
```

## Piecewise model

```{r segmented,echo=FALSE}
passing$season_star <- ifelse(passing$Season < 2015,0,passing$Season)
piecewise <- lme(Passes.Made ~ Season + season_star, random = ~1|Team,
                       data = passing,correlation = corAR1())
format_nlme_out(piecewise,varnames = c("Season","Change in Slope"),caption = NULL)
```

So, the overall passing rate doesn't appear to have changed since 2013.

# Assists

## Raw assist numbers by team and season

```{r raw assists,echo=FALSE,warning=FALSE,message=FALSE}
# Percentage of baskets assisted
all_seasons <- all_seasons %>%
  mutate(AST_perc = round(AST / FG,3)*100)
# Since merger
post_merger <- all_seasons[all_seasons$Season > 1976,]
post_merger <- post_merger %>% arrange(Team,Season)
# Raw assist number plot
ast_plot <- 
  ggplot(post_merger,aes(x = Season,y = AST,group = Team)) + 
  geom_line(aes(color = Team)) + theme_bw() + ylab("Assists") +
  scale_color_manual(values = primary_colors)
ast_plot
```

There were lockouts in 1998 and 2011.

## Percentage of FG Assisted

```{r pct ast,echo=FALSE,warning=FALSE,message=FALSE}
# Plot assist percentage by year and team
# All together
ast_perc_plot <- 
  ggplot(post_merger,aes(x = Season,y = AST_perc,group = Team)) + 
  geom_line(aes(color = Team)) + theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) +  
  ylab("Percentage of FG Assisted") + 
  scale_color_manual(values = primary_colors)
ast_perc_plot
```

## Steph Curry and John Stockton

```{r gsw uta,echo=FALSE}
# Facet wrap
ast_perc_plot <- 
  ggplot(post_merger[post_merger$Team %in% c("GSW","UTA"),],
         aes(x = Season,y = AST_perc,group = Team)) + 
  geom_line(aes(color = Team)) + theme_bw() + 
  theme(axis.text.x = element_text(angle = 90)) +  
  ylab("Percentage of FG Assisted") + 
  scale_color_manual(values = primary_colors)
ast_perc_plot +
  facet_wrap(~Team) +
  theme(legend.position = "none")
```

## Do assists help you win? 

Modeled winning percentage using normal theory mixed models.

```{r w hist,echo=FALSE}
post_merger$w_perc <- (as.numeric(as.character(post_merger$W)) / 
                         (as.numeric(as.character(post_merger$W)) + 
                            as.numeric(as.character(post_merger$L))))*100
hist(post_merger$w_perc,xlab = "Win %",main = "Histogram of Win %")
```

## Win model selection

```{r echo=FALSE}
mod <- lme(w_perc ~ AST_perc,random = ~1|Team,data = post_merger)
mod_s <- lme(w_perc ~ AST_perc,random = ~Season|Team,data = post_merger)
mod_ar1 <- lme(w_perc ~ AST_perc,random = ~1|Team,data = post_merger,
               correlation = corAR1())
mod_ar1_s <- lme(w_perc ~ AST_perc,random = ~Season|Team,data = post_merger,
               correlation = corAR1())
aics <- as.data.frame(AIC(mod,mod_s,mod_ar1,mod_ar1_s))
rownames(aics) <- c("RI Only","RI and RS","RI and AR(1)","RI, RS, and AR(1)")
kable(aics)
```

Compared the same general model types as for the passing model.

## Win model results

```{r W selection,echo=FALSE}
format_nlme_out(mod_ar1,varnames = "% of FG Assisted",caption = NULL)
```

An increase of 10 points in percentage of FGs assisted can lead to an approximately 5 point increase in winning percentage (or about 4 games) on the season. 

## Wins by percentage assisted

```{r w plot,echo=FALSE}
ggplot(post_merger,aes(x = AST_perc,y = w_perc,group = Team)) + 
  geom_point(aes(color = Team)) + 
  xlim(40,80) + 
  geom_abline(aes(intercept = 12.609, slope = 0.612)) + 
  xlab("Percentage of FG Assisted") + ylab("Win %") + 
  scale_color_manual(values = primary_colors) + 
  theme_bw() +
  theme(legend.position = "none")
```

## Wins by percentage assisted

```{r w plot labels,echo=FALSE}
ggplot(post_merger,aes(x = AST_perc,y = w_perc,group = Team)) + 
  geom_point(aes(color = Team)) + 
  xlim(40,80) + 
  geom_abline(aes(intercept = 12.609, slope = 0.612)) + 
  geom_label_repel(data = subset(post_merger,AST_perc >70 | w_perc > 85),
                   aes(label = team_season),
                   segment.color = 'grey50',
                   direction = "y",
                   nudge_x = 5,
                   hjust = 0.5) + 
  xlab("Percentage of FG Assisted") + ylab("Win %") + 
  scale_color_manual(values = primary_colors) + 
  theme_bw() +
  theme(legend.position = "none")
```

## Conclusions

- Passing hasn't increased since 2013.

- Assists are helpful! But they won't necessarily win you a championship.

## Limitations

- Assists are a very subjective measure.
- Assists also depend on the shooting percentage of the player receiving a pass.
- These analyses were not adjusted for any defensive measures, which leaves out a big part of team success and winning percentage. 
- They were also not adjusted for opponent defensive measures.
- Passing data does not go back very far.
