---
title: "BIOS 7659 Homework 7"
author: "Tim Vigers"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,dpi = 600)
library(shinyMethyl)
library(minfi) 
library(bumphunter) 
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(arsenal)
```

# 1. DNA Methylation QC and Normalization (Illumina 450K)

Load the data:

```{r warning=FALSE,message=FALSE}
baseDir = "C:/Users/tim/Dropbox/Documents/School/Statistical Genomics/Homework Files/HW7/idats"
targets = read.metharray.sheet(baseDir)
rgSet = read.metharray.exp(targets = targets)
annotation(rgSet)
```

## a) Table 1

```{r results='asis'}
df = as.data.frame(pData(rgSet))
df = df[df$Status == "cancer",]
t1 = tableby(~ patient.age_at_initial_pathologic_diagnosis +
               patient.height + patient.weight + Sex +  patient.race,
             data = df)
summary(t1,labelTranslations = 
          list(patient.age_at_initial_pathologic_diagnosis = 
                 "Age at Diagnosis",
               patient.height = "Height",patient.weight = "Weight",
               patient.race = "Race"))
```

There are three unique subjects in this dataset, each with two samples (one primary tumor sample and one from normal solid tissue).

## b) Type I and II probes

```{r}
getManifest(rgSet)
```

There are 135,476 type I probes and 350,036 type II probes. Type I probes have two different sequences per CpG site, one for methylated and one for unmethylated CpGs. Type II probes use a two-color channel, which allows each probe to measure both methylated and unmethylated CpGs. As a result, type II probes take up half the physical space of type I probes. However, they have a lower dynamic range than type I probes, and are also more biased and less reproducible. 

## c) QC Plots

### By ID

```{r}
id = pData(rgSet)$id
densityPlot(rgSet,sampGroups = id)
densityBeanPlot(rgSet,sampGroups = id)
```

### By sample type

```{r}
stype = pData(rgSet)$sample_type
densityPlot(rgSet,sampGroups = stype)
densityBeanPlot(rgSet,sampGroups = stype)
```

### By sex

```{r}
sex = pData(rgSet)$Sex
densityPlot(rgSet,sampGroups = sex)
densityBeanPlot(rgSet,sampGroups = sex)
```

## d) Control probes

```{r}
controlStripPlot(rgSet,controls = "BISULFITE CONVERSION I")
controlStripPlot(rgSet,controls = "NEGATIVE")
```

## e) Detection p values

```{r}
# Count p values >= 0.05 per sample
detect = detectionP(rgSet)
colSums(detect >= 0.05)
# Row means
rmeans = rowMeans(detect)
```

Sample `r colnames(detect)[which.max(colSums(detect >= 0.05))]` has the most detection p values $\geq 0.05$ with `r max(colSums(detect >= 0.05))` (`r round(max(colSums(detect >= 0.05))/nrow(detect)*100,3)`%). Out of the `r nrow(detect)` probes, `r sum(rmeans >= 0.05)` have a mean p value $\geq 0.05$.

### Save the methylation signals

```{r}
mset <- preprocessRaw(rgSet)
msetSWAN <- preprocessSWAN(rgSet)
```

## f) Multidimensional scaling (MDS) plots

```{r}
mdsPlot(msetSWAN)
```
